Automatic Annotation of Tasks in Structured Code

### 2018-07-14 docs/ACM/paper.tex
Adding references.

\begin{figure}[t!]
\begin{center}
\includegraphics[width=1\columnwidth]{images/ex_book_filter}
\caption{The benefits of the OpenMP runtime environment.
Neither the runtime checks of Alves~\cite{Alves15} or Rus~\cite{Rus02}, nor
Whaley's context and flow sensitive alias analysis would be able to ensure the
correctness of the automatic parallelization of this program.}
\label{fig:ex_book_filter}
\end{center}
\end{figure}

@@

Definition~\ref{def:task} uses the concept of {\em memory region}.
Syntactically, memory regions are described by program variables and/or pointers
plus ranges of dereferenceable offsets.
Given two tasks: $T_1 = (G_1, M_{i1}, M_{o1})$ and $T_2 = (G_2, M_{i2}, M_{o2})$,
if $M_{o1} \cap M_{i2} \neq \emptyset$, then we say that $T_2$ {\em depends} on
$T_1$.
If a program $P$ is partitioned into a set of $n$ tasks, then this partition is
said to be {\em correct} if it abides by Definition~\ref{def:correctness}.

\begin{definition}[Correctness]
\label{def:correctness}
A set $T$ of $n$ tasks is a correct parallelization of a program $P$ if:
\begin{compactenum}
\item $T$ does not contain cyclic dependence relations;
\item the execution of the tasks in $T$ in any ordering determined by dependence
relations leads to the same results as the sequential execution of $P$.
\end{compactenum}
\end{definition}

@@

\paragraph{The annotation zoo.}
Our equipment of choice to exploit task parallelism in programs is OpenMP 4.0.
%This annotation system lets us indicate to compiler
%and runtime environment when and how tasks must be created.
We describe below which annotations we are using, and explain, informally, their
semantics.
Full overviews of the syntax and semantics of these annotations are publicly
available\footnote{For a quick overview, we refer the reader to the leaflet
``Summary of OpenMP 4.0 C/C++ Syntax", which is made available by the OpenMP
group.}; hence, we shall not dive into their details.
%
\begin{compactitem}
\item \textsf{parallel} (\textit{clauses}): forms a team of threads
which will execute the marked program region in parallel.
\item \textsf{single} (\textit{clauses}): specifies that a
program region must be executed by a single thread in the team.
\item \textsf{task} (\textit{clauses}): creates a new task.
\item \textsf{taskwait}: defines a synchronization point where
threads must wait for the completion of child tasks.
\end{compactitem}
%
Some pragmas are associated with lists of clauses.
There are several of these clauses available in OpenMP 4.0; however, we only use
the following:
\begin{compactitem}
\item \textsf{default([shared/private])}: indicates that variables are either
shared or replicated among tasks.
\item \textsf{firstprivate($v$)}: indicates that $v$ must be replicated among
tasks, and initialized with its value at the point where the annotation is
executed.
\item \textsf{untied}: if a task has this modifier, then its associated code can
be executed by more than one thread; e.g., threads might alternate execution due
to preemption and load balancing.
\item \textsf{depend}($\mathit{in}/\mathit{out}/\mathit{inout}$):
determine if data is read ($\mathit{in}$), written ($\mathit{out}$) or both
within a task region; thus, effectively setting dependences among tasks.
\item \textsf{if}({\em condition}): this clause defines the conditions
under which a task is allowed to create threads.
We use it to enforce the cost model.
\end{compactitem}

@@

\noindent
\textbf{Runtime Environment:}
We have implemented the techniques described in this paper in
LLVM 3.9~\cite{Lattner04}.
All our experiments were performed in a 64-bits 12-core Intel(R) Xeon(R) CPU
E5-2620 at 2.00GHz, with 32K of L1 cache, 256K of L2 cache, 15M of L3 cache and
15Gb of main memory.
We use OpenMP 4.5, from November of 2015.

@@

Many systems have been developed to extract task parallelism from sequential
programs (semi) automatically.
Examples include OSCAR
(Optimally SCheduled Advanced multiprocessoR), Multigrain Parallelizing
Compiler \cite{ishizaka:journal:2000, kasahara:iwlcpc:2000},  MAPS (MPSoC
Application Programming  Studio) \cite{castrillon:tii:2013, ceng:dac:2008}, and
DiscoPoP (Discovery of Potential Parallelism) \cite{discopop, li:jss:2016}.
Besides parallelization systems, tools like Paraver \cite{extrae, paraver},
Aftermath~\cite{drebes:hipeac:2014}, DAGvis \cite{huynh:wvpa:2015}, and TEMANEJO
\cite{brinkmann:parco:2011, brinkmann:journal:2013, temanejo} were  designed to
enable performance analysis and visualization of task-based programs.
Although these tools help the programmer in adapting code
to run in parallel, they are not fully automatic.
For instance, the work that is the closest to ours, in reach and effectiveness,
in our opinion, is the suite of techniques proposed by Ravishankar {\em et
al.}~\cite{Ravishankar14}.
The type of irregular loops that they handle is impressive; however,
the lack of the runtime support \`{a} la OpenMP 4.X still requires them to
modify code before parallelization.
In their words: ``{\em For all benchmarks and applications, all functions were 
inlined, and arrays of structures were converted to structures of arrays for use
with our prototype compiler.}"

>>>>

\begin{figure}[t!]
\begin{center}
\includegraphics[width=1\columnwidth]{images/ex_book_filter}
\caption{The benefits of the OpenMP runtime environment.
Neither the runtime checks of Alves~\cite{Alves15} or Rus~\cite{Rus02}, nor
Whaley's context and flow sensitive alias analysis~\cite{Whaley04} would be able to ensure the
correctness of the automatic parallelization of this program.}
\label{fig:ex_book_filter}
\end{center}
\end{figure}

@@

Definition~\ref{def:task} uses the concept of {\em memory region}.
Syntactically, memory regions are described by program variables and/or pointers
plus ranges of dereferenceable offsets.
Given two tasks: $T_1 = (G_1, M_{i1}, M_{o1})$ and $T_2 = (G_2, M_{i2}, M_{o2})$,
if $M_{o1} \cap M_{i2} \neq \emptyset$, then we say that $T_2$ {\em depends} on
$T_1$.
If a program $P$ is partitioned into a set $T$ of $n$ tasks, then this partition is
said to be {\em correct} if:
(i) $T$ does not contain cyclic dependence relations; and
(ii) the execution of the tasks in $T$ in any ordering determined by dependence
relations leads to the same results as the sequential execution of $P$.

@@

\paragraph{The annotation zoo.}
Our equipment of choice to exploit task parallelism in programs is OpenMP 4.0.
%This annotation system lets us indicate to compiler
%and runtime environment when and how tasks must be created.
We describe below which annotations we are using, and explain, informally, their
semantics.
Full overviews of the syntax and semantics of these annotations are publicly
available\footnote{For a quick overview, we refer the reader to the leaflet
``Summary of OpenMP 4.0 C/C++ Syntax", which is made available by the OpenMP
group.}; hence, we shall not dive into their details.
%
\begin{compactitem}
\item \textsf{parallel} (\textit{clauses}): forms a team of threads
which will execute the marked program region in parallel.
\item \textsf{single} (\textit{clauses}): specifies that a
program region must be executed by a single thread in the team.
\item \textsf{task} (\textit{clauses}): creates a new task.
\item \textsf{taskwait}: defines a synchronization point where
threads must wait for the completion of child tasks.
\end{compactitem}
%
Some pragmas are associated with lists of clauses.
There are several of these clauses available in OpenMP 4.0; however, we only use
the following:
\begin{compactitem}
\item \textsf{default(shared)}: indicates that variables are
shared by default among tasks.
\item \textsf{firstprivate($v$)}: indicates that $v$ must be replicated among
tasks, and initialized with its value at the point where the annotation is
executed.
\item \textsf{untied}: if a task has this modifier, then its associated code can
be executed by more than one thread; e.g., threads might alternate execution due
to preemption and load balancing.
\item \textsf{depend}($\mathit{in}/\mathit{out}/\mathit{inout}$):
determine if data is read ($\mathit{in}$), written ($\mathit{out}$) or both
within a task region; thus, effectively setting dependences among tasks.
\item \textsf{if}({\em condition}): this clause leads to the creation of an {\em undeferred}
task, which must be executed immediately.
We use it to enforce the cost model.
\item \textsf{final}({\em condition}) \textsf{mergeable}: this clause limits the creation of
tasks, and reuses the data environment of the father task.
We use it prune tasks considered too small, as discussed in Section~\ref{sub:ir}.
\end{compactitem}

@@

\noindent
\textbf{Runtime Environment:}
We have implemented the techniques described in this paper in
LLVM 3.9~\cite{Lattner04}.
All our experiments were performed in a 64-bits 12-core Intel(R) Xeon(R) CPU
E5-2620 at 2.00GHz, with 32K of L1 cache, 256K of L2 cache, 15M of L3 cache and
15Gb of main memory.
We use OpenMP 4.0, from November of 2015.

@@

Many systems have been developed to extract task parallelism from sequential
programs (semi) automatically.
Examples include OSCAR
(Optimally SCheduled Advanced multiprocessoR), Multigrain Parallelizing
Compiler \cite{ishizaka:journal:2000, kasahara:iwlcpc:2000},  MAPS (MPSoC
Application Programming  Studio) \cite{castrillon:tii:2013, ceng:dac:2008}, and
DiscoPoP (Discovery of Potential Parallelism) \cite{discopop, li:jss:2016}.
Besides parallelization systems, tools like Paraver \cite{extrae, paraver},
Aftermath~\cite{drebes:hipeac:2014}, DAGvis \cite{huynh:wvpa:2015}, and TEMANEJO
\cite{Brinkmann13} were  designed to
enable performance analysis and visualization of task-based programs.
Although these tools help the programmer in adapting code
to run in parallel, they are not fully automatic.
For instance, the work that is the closest to ours, in reach and effectiveness,
in our opinion, is the suite of techniques proposed by Ravishankar {\em et
al.}~\cite{Ravishankar14}.
The type of irregular loops that they handle is impressive; however,
the lack of the runtime support \`{a} la OpenMP 4.X still requires them to
modify code before parallelization.
In their words: ``{\em For all benchmarks and applications, all functions were 
inlined, and arrays of structures were converted to structures of arrays for use
with our prototype compiler.}"
